Solving the Poisson-equation in one dimension \\ \hrulefill\small{ FYS3150: Computational Physics }\hrulefill

### 2016-09-13 Project1/article/article.tex
Working on algorithms in article

\author[Sebastian G. Winther-Larsen]{Sebastian G. Winther-Larsen \\
  \href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}

@@

By defining $\tilde{\vb{b}}=h^2f_i$ one can rewrite equation \ref{eq:2approx} as $-v_{i+1}-2v_i+v{i-1}=h^2f_i$. If we ignore the end points, $i=0$ and $i=n+1$, this equation can be represented as a matrix equation.
\begin{equation}
\begin{bmatrix}
2 & -1 & 0 & 0 & \cdots & 0 & 0 & 0 \\
-1 & 2 & -1 & 0 & \cdots & 0 & 0 & 0 \\
0 & -1 & 2 & -1 & \cdots & 0 & 0 & 0 \\ 
& & \vdots &  & \ddots &  & \vdots & \\
0 & 0 & 0 & 0 & \cdots & -1 & 2 & -1 \\
0 & 0 & 0 & 0 & \cdots & 0 & -1 & 2 
\end{bmatrix}
\begin{bmatrix}
v_1 \\
v_2 \\
v_3 \\
\vdots \\
v_{n-1} \\
v_n
\end{bmatrix}=
\begin{bmatrix}
b_1 \\
b_2 \\
b_3 \\
\vdots \\
b_{n-1} \\
b_n
\end{bmatrix}
\end{equation}

>>>>

\author[Svalheim \& Winther-Larsen]{Trygve Leithe Svalheim \\
   Sebastian G. Winther-Larsen \\
  \href{https://github.com/gregwinther/FYS3150/}{\texttt{github.com/gregwinther}}}

@@

By defining $\vb{f}=h^2f_i$ one can rewrite equation
\ref{eq:2approx} as $-v_{i+1}-2v_i+v_{i-1}=h^2f_i$. If we ignore the
end points, $i=0$ and $i=n+1$, this equation can be represented as a
matrix equation.
\begin{equation}
A\vb{v}=\vb{f}
\end{equation}
\begin{equation}
\begin{bmatrix}
2 & -1 & 0 & 0 & \cdots & 0 & 0 & 0 \\
-1 & 2 & -1 & 0 & \cdots & 0 & 0 & 0 \\
0 & -1 & 2 & -1 & \cdots & 0 & 0 & 0 \\ 
& & \vdots &  & \ddots &  & \vdots & \\
0 & 0 & 0 & 0 & \cdots & -1 & 2 & -1 \\
0 & 0 & 0 & 0 & \cdots & 0 & -1 & 2 
\end{bmatrix}
\begin{bmatrix}
v_1 \\
v_2 \\
v_3 \\
\vdots \\
v_{n-1} \\
v_n
\end{bmatrix}=
\begin{bmatrix}
f_1 \\
f_2 \\
f_3 \\
\vdots \\
f_{n-1} \\
f_n
\end{bmatrix}
\end{equation}

\section{Algorithms}
Two main methods are implemented and compared. The first method is
gaussian elimination of the tridiagonal matrix $A$, also known as the
\emph{Thomas Algorithm} \cite{thomasalgo}. This is a simplified form of Gaussian
elimination that can ve used to solve tridiagonal systems of
equations. The method is improved upon in order to take into account
the fact that the matrix we are dealing with has the same numbers
along the diagonals. The second method is the LU-decomposistion
method.

\subsection{Tridiagonal Matrix Algorithm}
Our tridiagonal system can be represented by 
\begin{equation}
a_ix_{i-1} + b_ix_i + c_ix_{i+1} = \vb{f}
\end{equation}
with $a_i=-1,b_i=2$ and $c_i=-1$, except for $a_1=0$ and $c_n=0$. Row
reducing a matrix will reveal how the algorithm functions. Limiting
the problem to four dimensions for easier reading and to save the
rainforest\footnote{Writing out a general case will take up more paper
space}.

\begin{equation}
\left[
\begin{array}{cccc|c}
b_1 & c_1 & 0 & 0 & f_1 \\
a_2 & b_2 & c_2 & 0  & f_2 \\
0 & a_3 & b_3 & c_3 & f_3 \\
0 & 0 & a_4 & b_4 & f_4
\end{array}
\right] \sim
\left[
\begin{array}{cccc|c}
1 & c_1/b_1 & 0 & 0 & b_1/b_1 \\
0 & b_2-\frac{c_1}{b_1}a_2 & c_2 & 0  & f_2-\frac{f_1}{b_1}a_2 \\
0 & a_3 & b_3 & c_3 & f_3 \\
0 & 0 & a_4 & b_4 & f_4
\end{array}
\right]
\end{equation}
Now let $\beta_1=b_1$ and $\beta_2=b_2-\frac{c_1}{b_1}a_2$. As new
elements start to appear in vector $\vb{f}$, right to the vertical bar
in the augmented matrix, they are also relabeled to $\tilde{f}_i$. For
example $\tilde{f}_1=f_1/\beta_1$. One more iteration will reveal the
pattern of the algorithm.

\begin{align}
\left[
\begin{array}{cccc|c}
1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
0 & 1 & c_2/\beta_2 & 0  & (f_2-\tilde{f}a_2)/\beta_2 \\
0 & a_3 & b_3 & c_3 & f_3 \\
0 & 0 & a_4 & b_4 & f_4
\end{array}
\right] &=
\left[
\begin{array}{cccc|c}
1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
0 & a_3 & b_3 & c_3 & f_3 \\
0 & 0 & a_4 & b_4 & f_4
\end{array}
\right] \\ \sim 
\left[
\begin{array}{cccc|c}
1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
0 & 0 & b_3-\frac{c_2}{\beta_2}a_3 & c_3 & f_3-\tilde{f}_2a_3 \\
0 & 0 & a_4 & b_4 & f_4
\end{array}
\right] &\sim 
\left[
\begin{array}{cccc|c}
1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
0 & 0 & 1 & c_3/\beta_3 & (f_3-\tilde{f}_2a_3)/\beta_3 \\
0 & 0 & a_4 & b_4 & f_4
\end{array}\right] \\ 
\sim &\dots \sim
\left[
\begin{array}{cccc|c}
1 & c_1/\beta_1 & 0 & 0 & \tilde{f}_1 \\
0 & 1 & c_2/\beta_2 & 0  & \tilde{f}_2 \\
0 & 0 & 1 & c_3/\beta_3 & \tilde{f}_3 \\
0 & 0 & 0 & 1 & \tilde{f}_4
\end{array}
\right]
\end{align}

This pattern can be summarized quite elegantly by the following
difference equations.
\begin{equation}
\beta_i=b_i-\frac{c_{i-1}}{\beta_{i-1}}, \quad
\tilde{f}_i=(\beta_i-\tilde{f}_{i-1}a_i)/\beta_i, \quad i \in [2,n], \quad
\end{equation}

\begin{thebibliography}{10}
  \bibitem{thomasalgo}{Thomas, L.H. (1949), \emph{Elliptic
        Problems in Linear Differential Equations over a
        Network}. Watson Sci. Comput. Lab Report, Columbia University,
      New York.}
\end{thebibliography}
